
# Making A List of Heatmaps


A list of heatmaps can improve visualization of the correspondence between multiple data sources.
In this vignette, we will discuss configurations for making a list of heatmaps and you can
see more real-world examples in the [**Examples**](s9.examples.html) vignette as well as [in the 
supplementaries](http://jokergoo.github.io/supplementary/ComplexHeatmap-supplementary1-4/index.html) of [the ComplexHeatmap paper](http://bioinformatics.oxfordjournals.org/content/early/2016/05/20/bioinformatics.btw313.abstract).

## Heatmap concatenation

You can arrange more than one heatmaps on the plot from left to right. Actually, one single
heatmap is just a special case of the heatmap list of length one.

`Heatmap()` is actually a class constructor function for a single heatmap. If more than one heatmaps
are to be combined, users can append one heatmap to the other by `+` operator.

```{r heatmap_list_default, fig.width = 10}
library(ComplexHeatmap)

set.seed(123)
mat1 = matrix(rnorm(80, 2), 8, 10)
mat1 = rbind(mat1, matrix(rnorm(40, -2), 4, 10))
rownames(mat1) = paste0("R", 1:12)
colnames(mat1) = paste0("C", 1:10)

mat2 = matrix(runif(60, max = 3, min = 1), 6, 10)
mat2 = rbind(mat2, matrix(runif(60, max = 2, min = 0), 6, 10))
rownames(mat2) = paste0("R", 1:12)
colnames(mat2) = paste0("C", 1:10)

le = sample(letters[1:3], 12, replace = TRUE)
names(le) = paste0("R", 1:12)

ind = sample(12, 12)
mat1 = mat1[ind, ]
mat2 = mat2[ind, ]
le = le[ind]

ht1 = Heatmap(mat1, name = "rnorm")
ht2 = Heatmap(mat2, name = "runif")
ht3 = Heatmap(le, name = "letters")
class(ht1)

ht1 + ht2 + ht3
```

Under default mode, dendrograms from the second heatmap will be removed and row orders will be same as the first one.

The returned value of addition of two heatmaps is a `HeatmapList` object. Directly calling `ht_list` object
will call `draw()` method with default settings. With explicitly calling `draw()` method, you can have more controls
e.g. on the legend and titles.

```{r}
ht_list = ht1 + ht2 + ht3
class(ht_list)
```

You can append any number of heatmaps to the heatmap list. Also you can append a heatmap list to a heatmap list.

```{r, eval = FALSE}
ht1 + ht1 + ht1
ht1 + ht_list
ht_list + ht1
ht_list + ht_list
```

`NULL` can be added to the heatmap list. It would be convinient when users want to construct a heatmap list through a `for` loop.

```{r, eval = FALSE}
ht_list = NULL  ## Heatmap(...) + NULL gives you a HeatmapList object
for(s in sth) {
    ht_list = ht_list + Heatmap(...)
}
```

## Titles

A heatmap list also has titles which are independent to the heatmap titles.

```{r heatmap_list_title, fig.width = 10}
col_rnorm = colorRamp2(c(-3, 0, 3), c("green", "white", "red"))
col_runif = colorRamp2(c(0, 3), c("white", "orange"))
col_letters = c("a" = "pink", "b" = "purple", "c" = "blue")
ht1 = Heatmap(mat1, name = "rnorm", col = col_rnorm,
    row_title = "Heatmap 1", column_title = "Heatmap 1",
    show_row_names = FALSE)
ht2 = Heatmap(mat2, name = "runif", col = col_runif,
    row_title = "Heatmap 2", column_title = "Heatmap 2",
    show_row_names = FALSE)
ht3 = Heatmap(le, name = "letters", col = col_letters)
ht_list = ht1 + ht2 + ht3

draw(ht_list, row_title = "Three heatmaps, row title", row_title_gp = gpar(col = "red"),
    column_title = "Three heatmaps, column title", column_title_side = "bottom")
```

```{r heatmap_list_gap, fig.width = 10, fig.keep = "all"}
draw(ht_list, ht_gap = unit(1, "cm"))
draw(ht_list, ht_gap = unit(c(3, 6), "mm"))
```

## Size of heatmaps

The width for some (not all) heatmaps can be set to a fixed width.

```{r heatmap_list_size, fig.width = 10, fig.keep = "all"}
ht2 = Heatmap(mat2, name = "runif", col = col_runif,
    width = unit(4, "cm"), show_row_names = FALSE)
ht3 = Heatmap(le, name = "letters", col = col_letters, heatmap_body_width = unit(5, "mm"))
ht1 + ht2 + ht3
```

or the width can be set as relative values. Please not in this case, `width` for all heatmaps
should be set (relative width and fixed width can be mixed).

```{r heatmap_list_relative_size, fig.width = 10, fig.keep = "all"}
ht1 = Heatmap(mat1, name = "rnorm", col = col_rnorm,
    show_row_names = FALSE, heatmap_body_width = unit(4, "cm"))
ht2 = Heatmap(mat2, name = "runif", col = col_runif,
    show_row_names = FALSE, heatmap_body_width = unit(6, "cm"))
ht3 = Heatmap(le, name = "letters", col = col_letters, heatmap_body_width = unit(1, "cm"))
ht1 + ht2 + ht3
```

## Auto adjustment to the main heatmap

There are some automatic adjustment if more than one heatmaps are plotted. There should be a main heatmap
which by default is the first one. Some settings for the remaining heatmaps will be modified to the settings
in the main heatmap. The adjustment are:

- row clusters are removed.
- row titles are removed.
- if the main heatmap is split by rows, all remaining heatmaps will also be split by same levels as the main one.

The main heatmap can be specified by `main_heatmap` argument. The value can be a numeric index or the name of the heatmap
(of course, you need to set the heatmap name when you create the `Heatmap` object).

```{r heatmap_list_auto_adjust, fig.width = 10, fig.keep = "all"}
ht1 = Heatmap(mat1, name = "rnorm", col = col_rnorm,
    show_row_names = FALSE, km = 2)
ht2 = Heatmap(mat2, name = "runif", col = col_runif,
    show_row_names = FALSE)
ht3 = Heatmap(le, name = "letters", col = col_letters)
ht1 + ht2 + ht3

# note we changed the order of `ht1` and `ht2`
ht2 + ht1 + ht3

# here although `ht1` is the second heatmap, we specify `ht1` to be
# the main heatmap by explicitely setting `main_heatmap` argument
ht_list = ht2 + ht1 + ht3
draw(ht_list, main_heatmap = "rnorm")
```

If there is no row clustering in the main heatmap, all other heatmaps have no row clustering neither.

```{r heatmap_list_auto_adjust_no_row_cluster, fig.width = 10}
ht1 = Heatmap(mat1, name = "rnorm", col = col_rnorm, cluster_rows = FALSE,
    show_row_names = FALSE)
ht2 = Heatmap(mat2, name = "runif", col = col_runif,
    show_row_names = FALSE)
ht3 = Heatmap(le, name = "letters", col = col_letters)
ht1 + ht2 + ht3
```

## control row ... in draw() function

following parameters that affect rows can be set for main heatmap directly
in draw():

row_dend_side
row_sub_title_side
column_dend_side
column_sub_title_side
row_gap = NULL
cluster_rows
clustering_distance_rows
clustering_method_rows = NULL, 
row_dend_width
show_row_dend
row_dend_reorder
row_dend_gp
row_order 
row_km
row_split

## Change graphic parameters simultaneously

`ht_global_opt()` can set graphic parameters for dimension names and titles as global settings.

```{r, fig.width = 10}
ht_global_opt(heatmap_row_names_gp = gpar(fontface = "italic"), 
	          heatmap_column_names_gp = gpar(fontsize = 14))
ht1 = Heatmap(mat1, name = "ht1", column_title = "Heatmap 1")
ht2 = Heatmap(mat2, name = "ht2", column_title = "Heatmap 2")
ht1 + ht2
ht_global_opt(RESET = TRUE)
```

Following are global settings supported by `ht_global_opt()`. By this function, you can also control settings
for the legends.

```{r}
names(ht_global_opt())
```

## Retrieve orders and dendrograms

`row_order`, `column_order`, `row_dend` and `column_dend` can be used to retrieve corresponding information from
the heatmaps. The usage is straightforward by following example:

```{r}
ht_list = ht1 + ht2
row_order(ht_list)
column_order(ht_list)
row_dend(ht_list)
column_dend(ht_list)
```

If `ht_list` has not been drawn yet, calling these four functions would be a little bit slow if the matrix are huge.
But if `ht_list` is already drawn which means clustering is already applied to the matrix, it will be fast to retrieve 
these information.

```{r, eval = FALSE}
ht_list = draw(ht1 + ht2)
row_order(ht_list)
column_order(ht_list)
row_dend(ht_list)
column_dend(ht_list)
```

## Heatmap list with row annotations

Row annotations can be added to the heatmap list, 
please see [**Heatmap Annotation**](s4.heatmap_annotation.html) for more explanations.


## the main heatmap

## concatenate with annotations

## vertical align

## Session info

```{r}
sessionInfo()
```
